---
title: "Effects of Historical Famines on Diabetes Prevelance"
output: html_document
date: "2023-12-11"
---

## Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, include=FALSE, echo=FALSE}
library(readxl)
library(readr)
library(tidyverse)
library(gt)
library(openintro)
library(modelr)
library(knitr)
library(xtable)
library(lterdatasampler)
library(tidymodels)
library(gghighlight)
library(glue)
library(ggmosaic)
library(ggridges)
library(gridExtra)
library(infer)
library(janitor)
library(kableExtra)
library(maps)
library(patchwork)
library(scales)
library(skimr)
library(caret)
library(palmerpenguins)
library(survival)
library(waffle)
library(ggrepel)
library(ggpubr)
```


```{r data import, echo=FALSE, include=FALSE}
df_famines <- read_excel("data/wpf_famines.xlsx", sheet = "owd_wfp_merged")
df_HNP_data <- read_excel("data/HNP_StatsEXCEL.xlsx", sheet = "Data")
df_WB_country <- read_excel("data/HNP_StatsEXCEL.xlsx", sheet = "Country")
df_WB_series <- read_excel("data/HNP_StatsEXCEL.xlsx", sheet = "Series")
df_owd_diabetes <- read_csv("data/diabetes-prevalence.csv")
```


## Data preparation

```{r}
#get list of just countries from the WB dataset
country_codes = (df_WB_country %>% filter(!is.na(`Currency Unit`)))$`Country Code`
df_HNP_countries = df_HNP_data %>% filter(`Country Code` %in% country_codes)
df_owd_diab_countries = df_owd_diabetes %>% filter(Code %in% country_codes)
```

First, we'll need to summarise and aggregate the famine data for each country.

```{r}

#cut off date for old famines is 1960 with the assumption that a relationship would take 2 generations to show full effects, since genes can skip a generation

df_famines_agg <- df_famines %>% summarize (.by = wb_ccode,
                                            famine_count = n(),
                                            avg_fam_length = mean(famine_length),
                                            avg_fam_mortality = mean(ex_mort_mid),
                                            max_fam_mortality = max(ex_mort_mid),
                                            total_fam_mortality = sum(ex_mort_mid),
                                            old_famines = sum(start_year<1960)
                                            ) %>%
                                            mutate (fam_occured = TRUE,
                                                    old_fam_occured = old_famines>0)

#Merge with the OWD diabetes list
df_owd_diab_countries <- df_owd_diab_countries %>% mutate(wb_ccode = Code)
df_owd_diab_fams = merge(x=df_owd_diab_countries, y=df_famines_agg, by = "wb_ccode", all.x = T )
df_owd_diab_fams = df_owd_diab_fams %>% mutate(fam_occured = !is.na(fam_occured))

#only keep latest 2021 data
df_owd_diab_fams_2021 = df_owd_diab_fams %>% filter(Year==2021)
df_owd_diab_fams_2011 = df_owd_diab_fams %>% filter(Year==2011)
```

We are interested in certain indicator variables from the World Bank HNP dataset, and we will need to group by each variable and make a new DF of countries. We will only look at the latest figures.

NY.GNP.PCAP.CD - GNI per capita, Atlas method (current US$)
SH.DYN.NCOM.ZS - Mortality from CVD, cancer, diabetes or CRD between exact ages 30 and 70 (%)
SH.STA.OWAD.ZS - Prevalence of overweight (% of adults)
SH.STA.DIAB.ZS - Diabetes prevalence (% of population ages 20 to 79)

```{r}
PCAP_country = df_HNP_countries %>% 
  filter(`Indicator Code`=="NY.GNP.PCAP.CD") %>% 
  select(`Country Code`,`Country Name`, `2021`) %>% 
  rename(wb_ccode = `Country Code`, NY.GNP.PCAP.CD=`2021`)

NCOM_country = df_HNP_countries %>% 
  filter(`Indicator Code`=="SH.DYN.NCOM.ZS") %>% 
  select(`Country Code`,`Country Name`, `2019`) %>% 
  rename(wb_ccode = `Country Code`, SH.DYN.NCOM.ZS=`2019`)

OWAD_country = df_HNP_countries %>% 
  filter(`Indicator Code`=="SH.STA.OWAD.ZS") %>% 
  select(`Country Code`,`Country Name`, `2016`) %>% 
  rename(wb_ccode = `Country Code`, SH.STA.OWAD.ZS=`2016`)

DIAB_country = df_HNP_countries %>% 
  filter(`Indicator Code`=="SH.STA.DIAB.ZS") %>% 
  select(`Country Code`,`Country Name`, `2021`) %>% 
  rename(wb_ccode = `Country Code`, SH.STA.DIAB.ZS=`2021`)

df_indicators = cbind(PCAP_country, 
                      SH.STA.DIAB.ZS = DIAB_country$SH.STA.DIAB.ZS, 
                      SH.DYN.NCOM.ZS = NCOM_country$SH.DYN.NCOM.ZS, 
                      SH.STA.OWAD.ZS = OWAD_country$SH.STA.OWAD.ZS)

df_HNP_indicators = merge(x=df_indicators, y=df_famines_agg, by = "wb_ccode", all.x = T )

df_HNP_indicators = df_HNP_indicators %>% 
  filter(!is.na(NY.GNP.PCAP.CD)) %>%
  mutate(fam_occured = !is.na(fam_occured))

df_HNP_famines = df_HNP_indicators %>% filter(fam_occured)
```


## Analysis
```{r}
ggplot(data = df_owd_diab_fams, aes(y = `Diabetes prevalence (% of population ages 20 to 79)`, x = as.factor(Year), colour = fam_occured)) +
  geom_point() +
  labs(x = "Year", y = "Diabetes prevalence (% of population)", title = "Diabetes prevalence (% of population ages 20 to 79)", subtitle = "All countries") + scale_color_discrete("Did famine occur?") + theme_classic()

ggplot(data = df_owd_diab_fams_2021, aes(y = `Diabetes prevalence (% of population ages 20 to 79)`, x = fam_occured)) +
  geom_point() +
  labs(x = "Famine occurrance",
       y = "Diabetes prevalence (% of population ages 20 to 79)", title = "Diabetes prevalence (% of population ages 20 to 79)", subtitle = "All countries") + theme_classic()

ggplot(data = df_owd_diab_fams_2011, aes(y = `Diabetes prevalence (% of population ages 20 to 79)`, x = old_fam_occured)) +
  geom_point() +
  labs(x = "Old Famine occurrance",
       y = "Diabetes prevalence (% of population ages 20 to 79)")+ theme_classic()


df_owd_diab_fams_2021 %>% filter(fam_occured == TRUE) %>%
  ggplot(aes(y=`Diabetes prevalence (% of population ages 20 to 79)`, x=old_fam_occured, label=Entity)) + geom_point() + geom_text(hjust=0, vjust=0) +
  labs(x = "Pre-1960 famine occurrance", y = "Diabetes prevalence (% of population)", title = "Diabetes prevalence (% of population ages 20 to 79)", subtitle = "Only Countries with Famines") + theme_classic()


```

Using World Bank Data

ggplot(data = df_owd_diab_fams, aes(y = `Diabetes prevalence (% of population ages 20 to 79)`, x = as.factor(Year), colour = fam_occured)) +
  geom_point() +
  labs(x = "Year", y = "Diabetes prevalence (% of population)", title = "Diabetes prevalence (% of population ages 20 to 79)", subtitle = "All countries") + theme_classic()
  
```{r}
mod_dia =lm (SH.STA.DIAB.ZS ~ fam_occured, data = df_HNP_indicators)
summary(mod_dia)
lm (SH.STA.DIAB.ZS ~ old_fam_occured, data = df_HNP_famines) %>% summary()

lm (SH.DYN.NCOM.ZS ~ fam_occured, data = df_HNP_indicators) %>% summary()
lm (SH.DYN.NCOM.ZS ~ old_fam_occured, data = df_HNP_famines) %>% summary()

lm (SH.STA.OWAD.ZS ~ fam_occured,  data = df_HNP_indicators) %>% summary()
lm (SH.STA.OWAD.ZS ~ old_fam_occured, data = df_HNP_famines) %>% summary()

mod_dia_gni = lm (SH.STA.DIAB.ZS ~ fam_occured + NY.GNP.PCAP.CD, data = df_HNP_indicators)
summary(mod_dia_gni)

mod_dia_gni_oldf =lm (SH.STA.DIAB.ZS ~ old_fam_occured + NY.GNP.PCAP.CD, data = df_HNP_famines)
summary(mod_dia_gni_oldf)

mod_dia_gni_owad = lm (SH.STA.DIAB.ZS ~ fam_occured + NY.GNP.PCAP.CD + SH.STA.OWAD.ZS, data = df_HNP_indicators)
summary(mod_dia_gni_owad)

mod_owad_fam_gni = lm (SH.STA.OWAD.ZS ~ fam_occured + NY.GNP.PCAP.CD, data = df_HNP_indicators)
summary(mod_owad_fam_gni)

df_HNP_indicators %>% 
  ggplot(aes(x = NY.GNP.PCAP.CD, y = SH.STA.OWAD.ZS, color = fam_occured)) +
  geom_point() +
  geom_line(data = augment(mod_owad_fam_gni), aes(y = .fitted, color = fam_occured)) + 
  labs(x = "GNI Per Capita (2021)",
       y = "Obesity prevelance (% adult population)") +
  scale_colour_discrete("Famine occurrance")

df_HNP_indicators %>% 
  ggplot(aes(x = NY.GNP.PCAP.CD, y = SH.STA.DIAB.ZS, color = fam_occured)) +
  geom_point() +
  geom_line(data = augment(mod_dia_gni), aes(y = .fitted, color = fam_occured)) + 
  labs(x = "GNI Per Capita (2021)",
       y = "Diabetes prevelance (% population) in 20-79 age group") +
  scale_colour_discrete("Famine occurrance")

df_HNP_famines %>% 
  ggplot(aes(x = NY.GNP.PCAP.CD, y = SH.STA.DIAB.ZS, color = old_fam_occured)) +
  geom_point() +
  geom_line(data = augment(mod_dia_gni_oldf), aes(y = .fitted, color = old_fam_occured)) + 
  labs(x = "GNI Per Capita (2021)",
       y = "Diabetes prevelance (% population) in 20-79 age group") +
  scale_colour_discrete("Old Famine occurrance")
```